---
title: "Bayesian threshold multiple-systems estimation"
author: "Bernard W. Silverman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian threshold multiple-systems estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(MSEBayesDev)
```

# Introduction

This vignette describes the Bayesian-threshold approach to multiple-systems
estimation developed in Silverman (2020), together with its implementation in
`estimate_population_bayesthresh()`.

Only the aspects specific to Bayesian thresholding are considered here.
General background on multiple-systems estimation and Poisson log-linear
models is given elsewhere in the `MultipleSystemsEstimation` package.

The basic Bayesian-threshold procedure described here follows Silverman
(2020). The extension to three-list interactions, and the use of the
Fienberg-Rinaldo existence check within this Bayesian-threshold
implementation, were not part of the original paper and are additions in
the present implementation.

# The Bayesian-threshold procedure

## Bayesian thresholding

In multiple-systems estimation, the choice of interaction terms in the
log-linear model can have a substantial effect on the estimate of the
unobserved population. Bayesian thresholding provides a way of selecting
these interactions using a posterior distribution rather than repeated
significance testing.

Initially, consider models which only have main effects and two-list
interactions. The approach begins with a model containing all possible
interactions, fits that model using MCMC, and then removes interactions for
which the posterior evidence of a non-zero effect is weak. The reduced model
is then fitted again to obtain the final population estimate and posterior
intervals.

The basic procedure therefore has two stages. First, a prior is chosen which
leaves the intercept and main effects essentially unconstrained, while possibly
allowing the interaction parameters to be shrunk towards zero.
For a model with intercept \(\mu\), main effects \(\alpha_i\), and interaction
parameters \(\beta_{ij}\), the prior assumes that the \(\mu\), \(\alpha_i\),
and \(\beta_{ij}\) are mutually independent. The intercept and main effects
are given diffuse priors while the interaction parameters are either given
Gaussian priors centred at zero with variance `prior_variance`, or themselves
have diffuse improper uniform priors on \((-\infty,\infty)\).

For each interaction parameter, the strength of evidence that it differs from
zero is measured by

\[
\frac{|\operatorname{E}(\beta_{ij}\mid\text{data})|}
     {\operatorname{sd}(\beta_{ij}\mid\text{data})}.
\]

Interactions for which this quantity is below the specified threshold are removed, and the model is then fitted again using the retained interactions.

The default threshold is 2. Thus the procedure retains interactions for which the absolute posterior mean is at least two posterior standard deviations from zero.

The method can also be extended to allow three-list interactions while
preserving the hierarchical structure of the log-linear model. The two-list
interactions are thresholded first, as described above. A three-list
interaction is then eligible for consideration only if all three of its
constituent two-list interactions have been retained. The model containing the
retained two-list interactions together with all eligible three-list
interactions is fitted, and the three-list interactions are thresholded using
the same posterior mean to posterior standard deviation ratio. The resulting
hierarchical model is then fitted once more to obtain the final population
estimate and posterior intervals.

For example, an interaction between lists \(i\), \(j\), and \(k\) is eligible
for inclusion only if the interactions \(ij\), \(ik\), and \(jk\) have all
been retained at the first stage. At the second stage, only the eligible
three-list terms are thresholded; the two-list terms that survived the initial
thresholding are kept regardless of their new posterior means. This ensures
that the selected model remains hierarchical.

## MCMC estimation

Posterior sampling is carried out using `MCMCpack::MCMCpoisson()`, which applies Markov chain Monte Carlo to the Poisson log-linear model. The MCMC run produces posterior draws for the intercept, the main effects, and the interaction parameters in the model. These draws are used both for the thresholding step described above and, after the selected model has been refitted, for inference about population size.

In the log-linear formulation, the intercept determines the fitted mean of the
unobserved cell, corresponding to individuals appearing on none of the lists.
For each posterior draw of the intercept \(\mu\), the corresponding size of
the unobserved cell is

\[
\exp(\mu).
\]

Adding this quantity to the observed population gives a posterior draw for the total population size. The resulting collection of draws therefore gives the posterior distribution of the total population. By default, `estimate_population_bayesthresh()` reports the posterior median as the population estimate, together with selected posterior quantiles. The full set of posterior draws can also be returned if required.

The thresholding step favours a parsimonious interaction structure; see
Silverman (2020) for further discussion.

# A basic analysis

The Kosovo data provide a useful illustration because allowing three-list
interactions leads to a genuinely different selected model.

We first fit the Bayesian-threshold procedure using only two-list
interactions.

```{r kosovo-pairwise}
data(Kosovo, package = "MultipleSystemsEstimation")

estimate_population_bayesthresh(
    Kosovo,
    maxorder = 2
)
```

With the default proper prior, the retained pairwise interactions are

EXH:ABA, EXH:OSCE, ABA:OSCE, EXH:HRW, and OSCE:HRW.

The posterior median estimate of the total population is about 13,971. The
2.5% and 97.5% posterior quantiles are about 12,236 and 16,389 respectively.

We can now allow three-list interactions by setting `maxorder = 3`.
For this example we use a threshold of 2.5, rather than the default value of
2, in order to illustrate the thresholding of eligible three-list
interactions.

```{r kosovo-threefactor}

estimate_population_bayesthresh(
    Kosovo,
    maxorder = 3,
    threshold = 2.5
)
```

The same five pairwise interactions are retained. Two three-list interactions
are eligible for consideration:

EXH:ABA:OSCE and EXH:OSCE:HRW.

Their threshold statistics are about 4.26 and 2.45 respectively. At a
threshold of 2.5, `EXH:ABA:OSCE` is retained while `EXH:OSCE:HRW` is
thresholded out.

The posterior median estimate of the total population is about 10,314, with
2.5% and 97.5% posterior quantiles of about 9,032 and 12,095 respectively.

Thus, in this example, allowing three-list interactions changes the selected
model and has a substantial effect on the inferred population size.

### Improper priors

The preceding analysis used a proper normal prior for the interaction
parameters. The same thresholding procedure can also be used with an
improper flat prior. Some additional care is needed because
an interaction whose sufficient statistic is zero has its posterior
distribution concentrated at minus infinity. Such terms are accounted for
before the MCMC fit is carried out.

Effects whose posterior distribution is concentrated at minus infinity
are identified automatically and reported in the
`minus_infinite_estimated_effects` component of the returned object. Such
effects are still retained in the model, but are not included in
`retained_interactions`, which reports only retained interaction effects
estimated by MCMC. For example, with the five-list UK data:

```{r uk-improper-minus-infinity}
data(UKdat_5, package = "MultipleSystemsEstimation")

estimate_population_bayesthresh(
  UKdat_5,
  prior = "improper"
)
```

Here `LA:GP` is retained in the model with its effect estimated as minus
infinity, and is therefore reported in `minus_infinite_estimated_effects`
rather than in `retained_interactions`.

### Checking existence of the fitted model

For some datasets, there are log-linear models for which there is no maximum
likelihood solution even after allowing some parameters to take the value
`-Inf`. This can be checked for any particular model using the criterion of
Fienberg and Rinaldo (2012).

Although the method is Bayesian, the Fienberg-Rinaldo existence condition is
still relevant because the prior specification used here is not fully proper
in all components of the model. If the condition fails, the resulting
posterior distribution need not be well defined, and the MCMC output cannot
be relied upon. The package therefore checks the existence condition before
carrying out the relevant thresholding step.

- **Initial pairwise fit.** If the model with all pairwise effects fails the
  Fienberg-Rinaldo check, the initial model fit will fail and it will not be
  possible to threshold the pairwise effects. The package returns an error
  message because a different approach altogether is needed.

- **Third-order fit.** In the `maxorder = 3` procedure, it is possible for
  the initial thresholding to work successfully, but for the model with all
  retained second-order and all eligible third-order interactions to fail the
  check. In that case, the completed order-2 analysis is returned instead of
  proceeding with the proposed order-3 fit.

### Examples of failure of the existence check

The two possible failures can be illustrated with the `Korea` and `Western`
datasets.

For `Korea`, the model containing all pairwise interactions fails the
Fienberg-Rinaldo check. The procedure therefore stops before the pairwise
effects can be thresholded.

```{r korea-existence, error=TRUE}
data(Korea, package = "MultipleSystemsEstimation")

estimate_population_bayesthresh(
  Korea
)
```

The second possibility is illustrated by `Western`. To make the issue visible
in a simple example, we deliberately use the unusually low value
`threshold = 0.2`. The initial pairwise fit and thresholding are then
completed successfully, but the resulting model containing the retained
pairwise interactions and all eligible third-order interactions fails the
Fienberg-Rinaldo check.

```{r western-existence}
data(Western, package = "MultipleSystemsEstimation")

estimate_population_bayesthresh(
  Western,
  maxorder = 3,
  threshold = 0.2
)
```

In this case, the procedure does not attempt the third-order fit and returns
the completed order-2 analysis, together with the eligible triples in the
proposed third-order extension.

# MCMC settings and posterior output

The principal MCMC settings can be controlled through arguments passed to
`MCMCpack::MCMCpoisson()`, including `burnin`, `mcmc`, `thin`, `tune`, and
`seed`. The defaults are intended to be suitable for routine use, although
longer runs may be appropriate for a substantive final analysis.
`MCMCpack::MCMCpoisson()` uses a reproducible default seed when `seed` is not
specified. A different seed can be supplied through `...` if a different MCMC
run is required.

By default, the full posterior sample of total population size is not returned.
It can be retained, when required, by setting `return_posterior = TRUE`.

# Summary

Bayesian thresholding provides a parsimonious way to select interaction effects
using posterior evidence. The implementation allows either proper or improper
priors for interaction effects, can extend the thresholding procedure
hierarchically to three-list interactions, and checks the relevant models
against the Fienberg-Rinaldo existence criterion. The returned object reports
the selected interaction structure and posterior inference for total population
size, with the full posterior sample available optionally.

# References

Silverman, B. W. (2020).
*Multiple-systems analysis for the quantification of modern slavery:
classical and Bayesian approaches*.
Journal of the Royal Statistical Society: Series A, **183**, 691--736.

Fienberg, S. E. and Rinaldo, A. (2012).
*Maximum likelihood estimation in log-linear models*.
The Annals of Statistics, **40**, 996--1023.

Martin, A. D., Quinn, K. M. and Park, J. H. (2011).
*MCMCpack: Markov Chain Monte Carlo in R*.
Journal of Statistical Software, **42**(9), 1--21.
